Approximation as Differentiation

### 2025-07-02 main/related-work.tex
Initial pass over AD.

>>>>

Automatic differentiation (AD), discussed in \secref{first-order:autodiff}, is the idea of computing
derivatives of functions expressed as programs by systematically applying the chain rule. The observation that
these derivative computations could be interleaved with the evaluation of the original program is due to
\citet{linnainmaa76}, who showed how the forward derivative $\pushf{f}_x$ of $f$ at a point $x$ could be
computed alongside $f(x)$ in a single pass, dramatically improving the efficiency of derivative evaluation
over symbolic or numerical differentiation. This insight became the foundation of forward-mode AD, which
underpins many optimisation and scientific computing tools, including JAX~\cite{jax2018github}.

More recent approaches to automatic differentiation have emphasised its semantic foundations.
\citet{elliott18} proposed a categorical model of AD that interprets programs not only as functions enriched
with their derivatives, giving a compositional account of differentiation based on duality and linear maps.
This line of work connects AD to denotational semantics, abstracting over operational concerns and enabling
principled extensions to differentiable programming languages. Vákár and
collaborators~\cite{vákár22,nunes2023} developed the CHAD framework, using Grothendieck constructions over
indexed categories to capture both values and their tangents in a unified semantic structure. These semantic
perspectives shed light on the categorical structure of AD and guide the design of systems that generalise
beyond real analysis, including the application to data provenance and slicing explored in this paper.
